Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-04-20 VegOzone.tex
Deduce -> determine

We deduced the dominant species on-site at Svanhovd and from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations. The localization of parameters for coniferous trees is based on MM's Boreal Norway spruce, deciduous trees on MM's silver birch, and perennial grassland on MM's perennial grassland for Central Europe. For a comprehensive list of MM parameters, consult Supplement~Table~S1. In the following, we suggest localizations of key environmental parameters to a subarctic climate based on meteorological data to assess over-/underestimation of ozone damage by using MM Boreal parameterizations.

@@

For deciduous trees (Fig.~\ref{fig:pody_rel}a) all $\mathrm{POD_1}$ simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Deciduous trees display the largest variation of ozone uptake among all PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light parameterization is pronounced. With MM GS, $\mathrm{POD_1}$ values were higher in 2018 than in 2019, throughout our analyses. However, taking the bespoke GS into account, this is reversed for the subarctic temperature acclimation. This can be explained by more favorable growing conditions in 2019. Low soil water potential reduced the $\mathrm{POD_1}$ values in 2018 in some cases when using the MM growing season, but did not or only slightly when using bespoke GS. The temperature acclimation-dependent difference in ozone uptake deduced for bespoke GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of $f_\mathrm{light}$ amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For coniferous trees (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). The MM Boreal Norway spruce parameterization was already well acclimated to the prevailing temperatures and therefore differs the least from our bespoke parameterizations. Taking the bespoke GS and canopy height into account, we find a higher ozone uptake in 2019 than in 2018. The overall difference deduced from bespoke GS for subarctic and cold parameterizations with respect to MM range between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of $f_\mathrm{light}$ and amounts to $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For perennial grassland (Fig.~\ref{fig:pody_rel}c), all simulations with bespoke GS stay below the CL. Perennial grassland shows the smallest ozone uptake, but a similarly large response to the temperature parameterizations as deciduous trees. Again, we find a reversal for predicted ozone risk in $\mathrm{POD_1}$ between 2018 and 2019 for the subarctic type. Perennial grassland shows the lowest sensitivity to the light threshold and SWP is only relevant without bespoke GS. The overall difference deduced from bespoke GS for subarctic and cold types with respect to MM ranges between $(2.2-3.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(4.4-6.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The lower maximum temperature tolerance in the subarctic compared to the cold temperature parameterization probably caused the lower $\mathrm{POD_1}$ values calculated for 2018 in subarctic compared to cold climate, whereas this was not seen for 2019. The maximum deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of $f_\mathrm{light}$ amount to $^{+0.3}_{-0.2}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+0.3}_{-0.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

The climatology for global irradiance ($Q_0$, not shown) approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). We deduce the corresponding highest $Q_0$ to $800\,\unit{W\,m^{-2}}$. The maximum of the lowest $Q_0$ (including night) amounts to $200\,\unit{W\,m^{-2}}$ indicating that a considerable amount of light is available under midnight sun conditions.

@@

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We deduced the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed an one-month latency period after snowmelt that is supported by observational data from Rovaniemi (Finland) \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season that is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

@@

In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures than for central European species. To this end, we computed a PDF from observed temperature at Svanhovd (May--September 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we deduced a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. Remark that these parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime.

>>>>

We determined the dominant species on-site at Svanhovd and from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations. The localization of parameters for coniferous trees is based on MM's Boreal Norway spruce, deciduous trees on MM's silver birch, and perennial grassland on MM's perennial grassland for Central Europe. For a comprehensive list of MM parameters, consult Supplement~Table~S1. In the following, we suggest localizations of key environmental parameters to a subarctic climate based on meteorological data to assess over-/underestimation of ozone damage by using MM Boreal parameterizations.

@@

For deciduous trees (Fig.~\ref{fig:pody_rel}a) all $\mathrm{POD_1}$ simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Deciduous trees display the largest variation of ozone uptake among all PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light parameterization is pronounced. With MM GS, $\mathrm{POD_1}$ values were higher in 2018 than in 2019, throughout our analyses. However, taking the bespoke GS into account, this is reversed for the subarctic temperature acclimation. This can be explained by more favorable growing conditions in 2019. Low soil water potential reduced the $\mathrm{POD_1}$ values in 2018 in some cases when using the MM growing season, but did not or only slightly when using bespoke GS. The temperature acclimation-dependent difference in ozone uptake determined for bespoke GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum deviations determined from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of $f_\mathrm{light}$ amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For coniferous trees (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). The MM Boreal Norway spruce parameterization was already well acclimated to the prevailing temperatures and therefore differs the least from our bespoke parameterizations. Taking the bespoke GS and canopy height into account, we find a higher ozone uptake in 2019 than in 2018. The overall difference determined from bespoke GS for subarctic and cold parameterizations with respect to MM range between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum deviations determined from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of $f_\mathrm{light}$ and amounts to $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For perennial grassland (Fig.~\ref{fig:pody_rel}c), all simulations with bespoke GS stay below the CL. Perennial grassland shows the smallest ozone uptake, but a similarly large response to the temperature parameterizations as deciduous trees. Again, we find a reversal for predicted ozone risk in $\mathrm{POD_1}$ between 2018 and 2019 for the subarctic type. Perennial grassland shows the lowest sensitivity to the light threshold and SWP is only relevant without bespoke GS. The overall difference determined from bespoke GS for subarctic and cold types with respect to MM ranges between $(2.2-3.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(4.4-6.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The lower maximum temperature tolerance in the subarctic compared to the cold temperature parameterization probably caused the lower $\mathrm{POD_1}$ values calculated for 2018 in subarctic compared to cold climate, whereas this was not seen for 2019. The maximum deviations determined from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of $f_\mathrm{light}$ amount to $^{+0.3}_{-0.2}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+0.3}_{-0.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

The climatology for global irradiance ($Q_0$, not shown) approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). We determine the corresponding highest $Q_0$ to $800\,\unit{W\,m^{-2}}$. The maximum of the lowest $Q_0$ (including night) amounts to $200\,\unit{W\,m^{-2}}$ indicating that a considerable amount of light is available under midnight sun conditions.

@@

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We determined the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed an one-month latency period after snowmelt that is supported by observational data from Rovaniemi (Finland) \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season that is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

@@

In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures than for central European species. To this end, we computed a PDF from observed temperature at Svanhovd (May--September 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we determined a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. Remark that these parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime.
